# Health Services and Policy Applications

## Patient Satisfaction survey

In this chapter, we reproduce the analyses reported in the manuscript available at https://doi.org/10.1186/s12913-025-12596-x
.

The first step in any empirical analysis is data acquisition. For this exercise, the dataset was imported into R using an R Markdown workflow. The Excel file (.xlsx), named DatasetNew, was stored on the desktop within a folder labeled Datasets and was read using the readxl package.

After importing the data, its structure was examined. The dataset contains 1,260 observations (rows) and 23 variables (columns).


```{r}
library(readxl)
library(dplyr) # Must load dplyr to use select()

DatasetNew <- read_excel("/Users/corneille/Desktop/Datasets/DatasetNew.xlsx")
# Exclude the unwanted column named '...4'
DatasetNew <- DatasetNew %>%
  select(-`...3`)

#Explore number of rows and # columns
dim(DatasetNew)
#look at structure
str(DatasetNew)
```

The second step is to check data quality. Here, I evaluated the percent of missingness for each variable.

```{r}
library(DataExplorer)

# Visualization
plot_missing (DatasetNew)
```



 
Next, I created a new data frame restricted to complete cases, retaining only observations with no missing values in the specified variables.
 
  
```{r}
# Load necessary packages
library(dplyr)
library(tidyr)

# 1. Define the set of variables to keep and check for completeness
varset <-

 c(

 'Province', 'visits', 'AgeCategory',
'Hospitalgrp', 'location', 'Gender', 'entrance', 'triage', 'medical doctor', 'laboratory', 'dispensing pharmacy', 'radiolohgy','rights respected?', 'waiting' , 'cashier'

 ) 

library(dplyr)

analytic.data <- DatasetNew %>%
  select(all_of(varset)) %>%      # keep only the variables in varset
  filter(complete.cases(.))       # keep only complete cases

dim(analytic.data)
```
 
Prior to regression modeling, data harmonization was performed to ensure consistency and comparability across variables. This included recoding categorical variables, aligning variable formats, and verifying reference categories, thereby ensuring that all variables were appropriately structured for regression analysis.
 
```{r}

#Recode Yes/No to 1/0

yes_no_vars <- c(
  "entrance",
  "triage",
  "medical doctor",
  "laboratory",
  "dispensing pharmacy",
  "radiolohgy",
  "rights respected?",
  "waiting",
  "cashier"
)

analytic.data <- analytic.data %>%
  mutate(across(all_of(yes_no_vars),
                ~ case_when(
                  . == "Yes" ~ 1,
                  . == "No"  ~ 0,
                  TRUE       ~ NA_real_
                )))

analytic.data <- analytic.data %>%
  mutate(
    Province = recode(Province,
                      "Eastern"      = 1,
                      "Kigali City"  = 2,
                      "Northern"     = 3,
                      "Southern"     = 4,
                      "Western"      = 5) |> as.numeric()
  )

analytic.data <- analytic.data %>%
  mutate(
    visits = recode(visits,
                    "new"        = 1,
                    "returning" = 2) |> as.numeric()
  )

analytic.data <- analytic.data %>%
  mutate(
    AgeCategory = recode(AgeCategory,
                         "young" = 1,
                         "Adult" = 2) |> as.numeric()
  )


analytic.data <- analytic.data %>%
  mutate(
    Hospitalgrp = recode(Hospitalgrp,
                         "District"     = 1,
                         "higherRanked" = 2) |> as.numeric()
  )

analytic.data <- analytic.data %>%
  mutate(
    location = recode(location,
                      "rural" = 1,
                      "urban" = 2) |> as.numeric()
  )

analytic.data <- analytic.data %>%
  mutate(
    Gender = recode(Gender,
                    "Male"   = 1,
                    "Female" = 2) |> as.numeric()
  )


```
 
 
 
Regression Analysis

The primary independent variable included in the regression model was location, which was specified according to the predefined categories in the dataset.

```{r}

###location#

#crude analysis

MOD1 <- glm(waiting ~ location, family = "binomial", data = analytic.data)
summary(MOD1)

# Odds ratios
exp(coef(MOD1))

exp(confint(MOD1))     

#Adjusted model


```

                        