---
title: "Statistical Methods for Survey, Epidemiology, Health Services and Policy Research"
author: "Corneille Killy Ntihabose"
date: "`r Sys.Date()`"
output: pdf_document
documentclass: book
bibliography:
- book.bib
- packages.bib
description: "This is a book on statistical and epidemiological methods applied to
  \nsurvey analysis, health services research, and policy evaluation.\n"
link-citations: true
github-repo: "ckntihabose-coder/Statistical-Methods-for-survey-Epidemiology-health-services-and-policy-research"
site: bookdown::bookdown_site
---

# About this book

This book is under development. It will eventually cover:

-   Survey methods for complex data (e.g., NHANES, DHS)
-   Regression models in epidemiology
-   Interrupted time series for policy evaluation
-   Health services and policy applications (drones, oxygen systems, etc.)

```{r, eval=FALSE}
bookdown::render_book()
```

To render this example to PDF as a `bookdown::pdf_book`, you'll need to install XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.org/tinytex/>.

## Preview book

As you work, you may start a local server to live preview this HTML book. This preview will update as you edit the book when you save individual .Rmd files. You can start the server in a work session by using the RStudio add-in "Preview book", or from the R console:

```{r eval=FALSE}
bookdown::serve_book()
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
