---
title: "Statistical Methods for Survey, Epidemiology, Health Services and Policy Research"
author: "Corneille Killy Ntihabose"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
description: |
  This is a book on statistical and epidemiological methods applied to 
  survey analysis, health services research, and policy evaluation.
link-citations: yes
github-repo: ckntihabose-coder/Statistical-Methods-for-survey-Epidemiology-health-services-and-policy-research
---

# About this book

Placeholder


## Preview book

<!--chapter:end:index.Rmd-->

# Hello bookdown 

All chapters start with a first-level heading followed by your chapter title, like the line above. There should be only one first-level heading (`#`) per .Rmd file.

## A section

All chapter sections start with a second-level (`##`) or higher heading followed by your section title, like the sections above and below here. You can have as many as you want within a chapter.

### An unnumbered section {-}

Chapters and sections are numbered by default. To un-number a heading, add a `{.unnumbered}` or the shorter `{-}` at the end of the heading, like in this section.

<!--chapter:end:01-intro.Rmd-->

# Cross-references {#cross}

Cross-references make it easier for your readers to find and link to elements in your book.

## Chapters and sub-chapters

There are two steps to cross-reference any heading:

1. Label the heading: `# Hello world {#nice-label}`. 
    - Leave the label off if you like the automated heading generated based on your heading title: for example, `# Hello world` = `# Hello world {#hello-world}`.
    - To label an un-numbered heading, use: `# Hello world {-#nice-label}` or `{# Hello world .unnumbered}`.

1. Next, reference the labeled heading anywhere in the text using `\@ref(nice-label)`; for example, please see Chapter \@ref(cross). 
    - If you prefer text as the link instead of a numbered reference use: [any text you want can go here](#cross).

## Captioned figures and tables

Figures and tables *with captions* can also be cross-referenced from elsewhere in your book using `\@ref(fig:chunk-label)` and `\@ref(tab:chunk-label)`, respectively.

See Figure \@ref(fig:nice-fig).

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center', fig.alt='Plot with connected points showing that vapor pressure of mercury increases exponentially as temperature increases.'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Don't miss Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(pressure, 10), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```


Data importing

The analysis utilizes data from five waves of the Rwanda Demographic and Health Surveys (DHS) spanning two decades (2000–2020). I imported the official SAS-formatted raw data files (.SAS7BDAT) for each cycle year into R using the haven package, verifying the file dimensions upon import for consistency.

The first step is to explore the file. The command was used to list the files that are contained within the specified directory path.

```{r}
list.files("/Users/corneille/Desktop/Mike lab/DHS/1992/RWBR21SD")
```


The library(haven) loads the haven package into the current R session.The haven package is essential because it provides functions (like read_sas) necessary to import data files saved in formats used by other statistical software, such as SAS (Statistical Analysis System), SPSS, and Stata.

The following lines read and load five different SAS data files (indicated by the .SAS7BDAT file extension) into the R environment, creating five separate data frames:

```{r}

library(haven)
data2000 <- read_sas("/Users/corneille/Desktop/Mike lab/DHS/2000/RWBR41SD/RWBR41FL.SAS7BDAT")

data2005 <- read_sas("/Users/corneille/Desktop/Mike lab/DHS/2005/RWBR53SD/RWBR53FL.SAS7BDAT")

data2010 <- read_sas("/Users/corneille/Desktop/Mike lab/DHS/2010/RWBR61SD/RWBR61FL.SAS7BDAT")

data2015 <- read_sas("/Users/corneille/Desktop/Mike lab/DHS/2015/RWBR70SD/RWBR70FL.SAS7BDAT")

data2020 <- read_sas("/Users/corneille/Desktop/Mike lab/DHS/2020/RWBR81SD/RWBR81FL.SAS7BDAT")

```



Finally, The five commands were executed to inspect the dimensions (number of rows and columns) of each of the newly loaded data frames. This is a crucial initial step to verify that the data was imported correctly and to understand the size of each dataset (rows = number of observations/survey participants; columns = number of variables/questions).
```{r}
dim(data2000)
dim(data2005)
dim(data2010)
dim(data2015)
dim(data2020)

```





<!--chapter:end:02-survey-methods.Rmd-->

# Parts

You can add parts to organize one or more book chapters together. Parts can be inserted at the top of an .Rmd file, before the first-level chapter heading in that same file. 

Add a numbered part: `# (PART) Act one {-}` (followed by `# A chapter`)

Add an unnumbered part: `# (PART\*) Act one {-}` (followed by `# A chapter`)

Add an appendix as a special kind of un-numbered part: `# (APPENDIX) Other stuff {-}` (followed by `# A chapter`). Chapters in an appendix are prepended with letters instead of numbers.




<!--chapter:end:03-epidemiology-methods.Rmd-->

# Footnotes and citations 

## Footnotes

Footnotes are put inside the square brackets after a caret `^[]`. Like this one ^[This is a footnote.]. 

## Citations

Reference items in your bibliography file(s) using `@key`.

For example, we are using the **bookdown** package [@R-bookdown] (check out the last code chunk in index.Rmd to see how this citation key was added) in this sample book, which was built on top of R Markdown and **knitr** [@xie2015] (this citation was added manually in an external file book.bib). 
Note that the `.bib` files need to be listed in the index.Rmd with the YAML `bibliography` key.

<!--bs4_book:start-->

The `bs4_book` theme makes footnotes appear inline when you click on them. In this example book, we added `csl: chicago-fullnote-bibliography.csl` to the `index.Rmd` YAML, and include the `.csl` file. To download a new style, we recommend: https://www.zotero.org/styles/

<!--bs4_book:end-->

The RStudio Visual Markdown Editor can also make it easier to insert citations: <https://rstudio.github.io/visual-markdown-editing/#/citations>

<!--chapter:end:04-health-services.Rmd-->


# Blocks

Placeholder


## Equations
## Theorems and proofs
## Callout blocks

<!--chapter:end:05-ITS-analysis.Rmd-->


# Sharing your book

Placeholder


## Publishing
## 404 pages
## Metadata for sharing

<!--chapter:end:06-NHANES-survival.Rmd-->

`r if (knitr::is_html_output()) '
# References {-}
'`

<!--chapter:end:07-DHS-stunting.Rmd-->

`r if (knitr::is_html_output()) '
# References {-}
'`

<!--chapter:end:08-missing-data.Rmd-->

`r if (knitr::is_html_output()) '
# References {-}
'`

<!--chapter:end:09-appendix.Rmd-->

